---
title: "Case Study - Bike Share"
author: "Andrea"
date: "2/6/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is a case study of a bike sharing company that wants his marketing department to create campaigns targeting his casual riders so that some of they could become members, customers of the annual membership.

## Business Task

The business task is understand the behavior of both types of customers: casual riders and members. Understand the differences in their use of the services so that the company can maximize the number of member (more profitable users for the company).


## Data sources

The data analyzed are the data trips of the twelve previous months. The data is public and has been made available by Motivate International Inc. under this [license](https://www.divvybikes.com/data-license-agreement).

All data is anonymous, there is no personal data from riders.

## Data cleaning

Load packages to be used.
```{r Load packages to be used}
library(tidyverse)
library(lubridate)
library(ggplot2)
```

Import data from 12 previous months from the working directory
```{r Import data from 12 previous months from the working directory}
X202004_divvy_tripdata <- read_csv("202004-divvy-tripdata.csv")
X202005_divvy_tripdata <- read_csv("202005-divvy-tripdata.csv")
X202006_divvy_tripdata <- read_csv("202006-divvy-tripdata.csv")
X202007_divvy_tripdata <- read_csv("202007-divvy-tripdata.csv")
X202008_divvy_tripdata <- read_csv("202008-divvy-tripdata.csv")
X202009_divvy_tripdata <- read_csv("202009-divvy-tripdata.csv")
X202010_divvy_tripdata <- read_csv("202010-divvy-tripdata.csv")
X202011_divvy_tripdata <- read_csv("202011-divvy-tripdata.csv")
X202012_divvy_tripdata <- read_csv("202012-divvy-tripdata.csv")
X202101_divvy_tripdata <- read_csv("202101-divvy-tripdata.csv")
X202102_divvy_tripdata <- read_csv("202102-divvy-tripdata.csv")
X202103_divvy_tripdata <- read_csv("202103-divvy-tripdata.csv")
X202104_divvy_tripdata <- read_csv("202104-divvy-tripdata.csv")
```
Mutate the format of start station and end station into character in the files that has a double format
```{r Mutate the format of start station and end station into character in the files that has a double format}
X202011_divvy_tripdata <-  mutate(X202011_divvy_tripdata, 
                                  start_station_id = as.character(start_station_id), 
                                  end_station_id = as.character(end_station_id))
X202010_divvy_tripdata <-  mutate(X202010_divvy_tripdata, 
                                  start_station_id = as.character(start_station_id), 
                                  end_station_id = as.character(end_station_id)) 
X202009_divvy_tripdata <-  mutate(X202009_divvy_tripdata, 
                                  start_station_id = as.character(start_station_id), 
                                  end_station_id = as.character(end_station_id)) 
X202008_divvy_tripdata <-  mutate(X202008_divvy_tripdata, 
                                  start_station_id = as.character(start_station_id), 
                                  end_station_id = as.character(end_station_id)) 
X202007_divvy_tripdata <-  mutate(X202007_divvy_tripdata, 
                                  start_station_id = as.character(start_station_id), 
                                  end_station_id = as.character(end_station_id)) 
X202006_divvy_tripdata <-  mutate(X202006_divvy_tripdata, 
                                  start_station_id = as.character(start_station_id), 
                                  end_station_id = as.character(end_station_id)) 
X202005_divvy_tripdata <-  mutate(X202005_divvy_tripdata, 
                                  start_station_id = as.character(start_station_id), 
                                  end_station_id = as.character(end_station_id)) 
X202004_divvy_tripdata <-  mutate(X202004_divvy_tripdata, 
                                  start_station_id = as.character(start_station_id), 
                                  end_station_id = as.character(end_station_id))
```

Join all data frames
```{r Join all data frames}
all_trips <- bind_rows(X202004_divvy_tripdata, X202005_divvy_tripdata, X202006_divvy_tripdata, 
                       X202007_divvy_tripdata, X202008_divvy_tripdata, X202009_divvy_tripdata, 
                       X202010_divvy_tripdata, X202011_divvy_tripdata, X202012_divvy_tripdata, 
                       X202101_divvy_tripdata, X202102_divvy_tripdata, X202103_divvy_tripdata, 
                       X202104_divvy_tripdata)
```

Inspect the dataset to see columns, structure, formats and summary
```{r Inspect the dataset to see columns, structure, formats and summary}
colnames(all_trips)  
nrow(all_trips)  
dim(all_trips) 
head(all_trips) 
str(all_trips) 
summary(all_trips) 
```

Add columns that list the date, month, day, and year of each ride so that later we can agregate by them
```{r Add columns that list the date, month, day, and year of each ride so that later we can agregate by them}
all_trips$date <- as.Date(all_trips$started_at)
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")

```
 
Add a ride_length calculation to all_trips (in seconds) and check all new fields
```{r Add a ride_length calculation to all_trips (in seconds)}
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)
all_trips %>% 
  select(date, month, day, year, day_of_week, ride_length)
```

Convert "ride_length" from Factor to numeric so we can run calculations on the data
```{r Convert "ride_length" from Factor to numeric so we can run calculations on the data}

all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)
```

Remove bad data (quality check trips and rows with ride_length negative) and create a new version
```{r Remove bad data (quality checks trips and rows with ride_length negative) and create a new version}
all_trips_v2 <- all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_length<0),]
```


Created a new data frame with all the duplicates from the last version to check that all have ride_id with value NA
```{r Created a new data frame with all the duplicates from the last version to check that all have ride_id with value NA}
duplicates <- data.frame(ride_id = all_trips_v2$ride_id , dup_indicator = duplicated(all_trips_v2$ride_id), 
                         started_at = all_trips_v2$started_at, ended_at = all_trips_v2$ended_at)
filter(duplicates, dup_indicator=="TRUE")
```

## Analysis

Descriptive analysis on ride_length (all figures in seconds)
```{r Descriptive analysis on ride_length (all figures in seconds)}
summary(all_trips_v2$ride_length)
```
Compare members and casual users
```{r Compare members and casual users}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)
```
See the average ride time and number of rides by each day for members vs casual users
```{r See the average ride time and number of rides by each day for members vs casual users}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = length)
```

Analyze ridership data by user type and weekday
```{r Analyze ridership data by user type and weekday}


all_trips_v2 %>% 
  group_by(member_casual, day_of_week) %>%  
  summarise(number_of_rides = n()		
            ,average_duration = mean(ride_length)) %>% 	
  arrange(member_casual, day_of_week)	
```

From this information you can conclude that:

* Casual users: use the service for leisure, taking long rides mostly on weekends.
* Members: use the service for work/school purposes, taking shorter rides mostly on weekdays.

## Visualizations

Number of rides by rider type and day of the week
```{r Number of rides by rider type}
all_trips_v2 %>% 
  group_by(member_casual, day_of_week) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  drop_na(day_of_week) %>% 
  arrange(member_casual, day_of_week)  %>% 
  ggplot(aes(x = day_of_week, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(title = "Number of rides by rider type and day of the week")
```

Average duration of rides by rider type and day of the week
```{r Average duration}
all_trips_v2 %>% 
  group_by(member_casual, day_of_week) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  drop_na(day_of_week) %>%
  arrange(member_casual, day_of_week)  %>% 
  ggplot(aes(x = day_of_week, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(title = "Average duration of rides by rider type and day of the week")
```
Create a summary csv for use in other software
```{r Create a csv for use in other software}
res <- all_trips_v2 %>% 
  group_by(member_casual, day_of_week) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  drop_na(day_of_week) %>%
  arrange(member_casual, day_of_week)
write.csv(res, file = "res.csv")
```


## Recomendations for the marketing team
### Where should their digital (social media, email and other) campaigns  focus

* Encourage casual riders to use the service regularly for work or school.
* Encourage causal riders to use the service for shorter rides (run errands, dates etc.) reminding of the benefits of using it instead of other means of transportation.
* Equate the biking service as other mean of transport so it can be used on a daily basis, not only for leisure.

